---
title: "HW1: Data Manipulation"
author: "Will Akin"
date: "Due: 02/22/26 (Submit BOTH RMD and HTML to Brightspace)"
---

# 0. Instructions (READ FIRST)

- Download the `HW1.Rmd` file from Brightspace.
- Move `HW1.Rmd` into our class R project folder.
- Open `HW1.Rmd` in RStudio.
- Fill in your answers directly in `HW1.Rmd`.
- Knit `HW1.Rmd` (this will create `HW1.html`).
- Submit **both** `HW1.Rmd` and `HW1.html` to Brightspace.
- Due **02/22/26**.

# 1. Business scenario

Suppose you work for an animal nutrition company that sells 4 feed formulas (Diet 1–4).

Your manager might ask: Which diet produces the **largest average weight gain** from day 0 to day 21, or other related questions.

You will try to answer these questions using tidy data manipulation and clear visuals.

# 2. Setup and data load with base R (don't change code in this section)

```{r,message = FALSE}
# CORE packages used in lecture for data manipulation
library(tidyverse)   # dplyr, tidyr, ggplot2, readr, tibble, etc.
library(lubridate)   # date tools (we will create a date column)
library(scales)      # formatting numbers in plots/tables
data("ChickWeight")
cw_raw <- as_tibble(ChickWeight)
```

# 3. Check data structure

**Instruction:**

- Use a function to check the head of `cw_raw`.

So, you are expected to see the **original variables** below:

- `weight`: chick body weight (grams)
- `Time`: time in days
- `Chick`: chick ID
- `Diet`: diet group (1–4)

```{r}
### Insert your code below
head(cw_raw)
```

# 4. Rename and type-clean

**Instruction:**  
Create `cw` with these changes:

- rename below variables
  - rename `Time` → `day` (hint: use the `rename()` function; if you are unsure about the parameters, run `?rename` in the Console to open the Help page in RStudio)
  - rename `Diet` → `diet`
  - rename `Chick` → `chick`
- mutate below variables
  - convert `chick` to character (so it behaves like an ID)
  - convert `diet` to factor
- arrange by `diet`, `chick`, `day`
- check the head of the `cw` data

```{r}
### Insert your code below
cw <- cw_raw %>% 
  rename(day = Time) %>% 
  rename(diet = Diet) %>% 
  rename(chick = Chick) %>% 
  mutate(chick = as.character(chick)) %>% 
  mutate(diet = as.factor(diet)) %>%
  arrange(diet, chick, day)

head(cw)
```

# 5. Compute average weight by diet and day

**Instruction:** Create `diet_curve` dataset:

- group by `diet` and `day`
- summarise the information below (drop the groups at the end)
  - compute new variables: `avg_weight` (mean), `sd_weight` (standard deviation), and `n`
- mutate `se = sd_weight / sqrt(n)`
- check the head of the `diet_curve` data

```{r}
### Insert your code below
diet_curve <- cw %>% 
  group_by(diet, day) %>% 
  summarise(
  avg_weight = mean(weight), 
  sd_weight = sd(weight),
  n = n(),
  .groups = "drop") %>% 
  mutate(se = sd_weight / sqrt(n))
head(diet_curve)
```

# 6. Plot growth curves based on `diet_curve`

**Instruction:** Make a line chart:

- x = `day`
- y = `avg_weight`
- color = `diet`
- line chart
- include points (hint: use `geom_point()`)
- y-axis scale with labels
- in `labs()`, include a clear title and axis labels
- use `theme_minimal()`

```{r}
### Insert your code below
diet_curve %>% 
  ggplot(aes(x = day, y = avg_weight)) +
  geom_line() +
  scale_y_continuous(labels = label_number()) +
  labs(
    x = "Day",
    y = "Avg Weight",
    title = "Avg Weight By Day"
  ) +
  theme_minimal()
```

# 7. Create a synthetic date

In lecture, we practiced creating date columns using `lubridate`. Here we create a synthetic calendar date starting at **2026-01-01**, using `cw` data at Section 4.

**Instruction:**

- define `start_date` as the date of **2026-01-01**.
- using pipe operator update `cw` dataset, and mutate the `date` variable using `start_date` and `day` (make sure `day` is converted to a `days()` format).
- check the head of the `cw` data again.

```{r}
### insert your code below
start_date <- as.Date("2026-01-01")

cw |>
  mutate(date = start_date + days(day))

head(cw)
```

# 8. Keep chicks with the full treatment period

**Instruction:** Some chicks have missing days; keep only chicks that have **both day 0 and day 21**.

- continue update the `cw` dataset.
- group by `chick`.
- filter to keep only chicks whose `day` values include both 0 and 21 (hint: use `all(c(0, 21) %in% day)` to return a single logical value).
- ungroup.

```{r}
### insert your code below
cw %>% 
  group_by(chick) %>% 
  filter(all(c(0, 21) %in% day)) %>% 
  ungroup()
```

# 9. Create a per-chick-diet summary table

**Instruction:** Create `chick_growth` with one row per chick and diet:

- name this summary dataset `chick_growth` based on `cw`.
- group by `chick` and `diet`.
- summarise two weight variables at day 0 and 21 (drop the groups at the end):
  - `w0` = weight at day 0 (hint: `new_variable = weight[condition_for_day_0]`)
  - `w21` = weight at day 21
- mutate two variables for gain and percent gain:
  - `gain_0_21 = w21 - w0`
  - `pct_gain_0_21 = (w21 / w0) - 1`
- arrange/order the `diet` and descending order for `gain_0_21`
- check the head of `chick_growth` with 10 observations.

```{r}
### insert your code below
chick_growth <- cw %>% 
  group_by(chick, diet) %>% 
  summarise(
    w0 = weight[day == 0],
    w21 = weight[day == 21],
    .groups = "drop"
  ) %>% 
  mutate(gain_0_21 = w21 - w0,
         pct_gain_0_21 = (w21 / w0) - 1) %>% 
  arrange(desc(gain_0_21))
head(chick_growth)
```

# 10. Diet-level KPI table

**Instruction:** Create `diet_kpis` with one row per diet based on `chick_growth`:

- name this summary dataset `diet_kpis` based on `chick_growth`.
- group by `diet`.
- summarise the information below (drop the groups at the end):
  - `n_chicks`: number of chicks per diet
  - `avg_gain`: mean of `gain_0_21`
  - `med_gain`: median of `gain_0_21`
  - `sd_gain`: standard deviation of `gain_0_21`
  - `avg_pct_gain`: mean of `pct_gain_0_21`
- sort by `avg_gain` (descending).
- check the head of `diet_kpis`.

```{r}
### insert your code below
diet_kpis <- chick_growth %>% 
  group_by(diet) %>% 
  summarise(
    n_chicks = n(),
    avg_gain = mean(gain_0_21),
    med_gain = median(gain_0_21),
    sd_gain = sd(gain_0_21),
    avg_pct_gain = mean(pct_gain_0_21),
    .groups = "drop") %>% 
    arrange(desc(avg_gain))
head(diet_kpis)
```

# 11. Written takeaway

In 2–4 sentences, conclude the data analysis above (write your paragraph below):
# Throughout this data analysis, I looked at the growth trend of chicks over 21 days.  Additionally, the chicks were segmented into 4 different diets, where effects on weight gain were analyzed.  The average weight gain for chicks on diet 1 = 136 , diet 2 = 174, diet 3 = 229.5, and diet 4 = 197.67  Diet 3 had the most significant weight gain.
























