---
title: "Lab6: European Stock Markets Dashboard (EuStockMarkets)"
format:
  dashboard:
    orientation: columns
---

```{r}
#| label: load-packages
#| message: false # hides code when rendered
library(tidyverse)
library(lubridate)
library(gt)
library(scales)
```

```{r}
#| label: load-data
#| message: false

# Built-in R dataset

data("EuStockMarkets")  # European stock data
```


```{r}
#| label: prep-data
#| message: false

# EuStockMarkets is a time-series (ts) matrix with 4 indices: DAX, SMI, CAC, FTSE

# Convert to tidy long data with a synthetic date index (business days sequence)

eu_tbl <- as.data.frame(EuStockMarkets) |>
tibble::rownames_to_column(var = "t") |>
mutate(t = as.integer(t))

# Create an approximate date sequence (dataset has 1860 observations; original is daily trading data)

start_date <- as.Date("1991-01-01")
eu <- eu_tbl |>
mutate(date = start_date + days(t - 1)) |>
select(date, DAX, SMI, CAC, FTSE) |>
pivot_longer(cols = c(DAX, SMI, CAC, FTSE), names_to = "index", values_to = "price") |>
arrange(index, date) |>
group_by(index) |>
mutate(
daily_return = price / lag(price) - 1,
month = floor_date(date, "month"),
year = year(date)
) |>
ungroup()

# Latest snapshot per index

latest <- eu |>
group_by(index) |>
filter(date == max(date)) |>
summarise(
latest_date = max(date),
latest_price = last(price),
.groups = "drop"
)

# Biggest gainers/losers by cumulative return (start to end)

cum_perf <- eu |>
group_by(index) |>
summarise(
start_price = first(price),
end_price = last(price),
cum_return = (end_price / start_price) - 1,
.groups = "drop"
) |>
arrange(desc(cum_return))

```


<br>

This dashboard uses the built-in R dataset EuStockMarkets. **Download the Quarto at [here](https://quarto.org/docs/get-started/) first, then you can run the Lab6.**

```{r}
#| label: split-indices
#| message: false
dax <- eu |> filter(index == "DAX")
ftse <- eu |> filter(index == "FTSE")
```

# 📈 DAX Dashboard

## Column - Prices and returns over time {width="20%"}


```{r}
#| label: valuebox-generall-1
#| message: false
top_index <- cum_perf |> slice_head(n = 1)
bottom_index <- cum_perf |> slice_tail(n = 1)

dax_latest <- latest |> filter(index == "DAX")
ftse_latest <- latest |> filter(index == "FTSE")

```

::: {.valuebox icon="graph-up-arrow" color="#2b8cbe"}
Best cumulative performer:

`{r} as.character(top_index$index)`

`{r} percent(top_index$cum_return, accuracy = 0.1)`
:::


::: {.valuebox icon="graph-down-arrow" color="#d95f0e"}
Worst cumulative performer:

`{r} as.character(bottom_index$index)`

`{r} percent(bottom_index$cum_return, accuracy = 0.1)`
:::

::: {.valuebox icon="calendar-event" color="#756bb1"}
Latest observation date:

`{r} as.character(max(latest$latest_date))`
:::

## Column - Prices and returns over time {width="45%"}

### Row - Price level over time {height="60%"}

```{r}
#| label: dax-price
#| title: DAX — Price level
#| fig-width: 10
#| fig-asp: 0.618 # Gold ratio
dax |>
ggplot(aes(x = date, y = price)) +
geom_line() +
scale_y_continuous(labels = label_number()) +
labs(x = "Date", y = "price") +
theme_minimal()


```

### Row - Monthly average returns {height="40%"}

::: {.card title="DAX — Monthly average returns"}
Average daily return within each month (approx).

```{r}
#| label: dax-monthly-returns
#| fig-width: 10
#| fig-asp: 0.3
dax |>
group_by(month) |>
summarise(avg_daily_return = mean(daily_return, na.rm = TRUE), .groups = "drop") |>
ggplot(aes(x = month, y = avg_daily_return)) +
geom_line() +
geom_point(size = 0.8) +
scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
labs(x = "Month", y = "Return") +
theme_minimal()
```
:::


## Column - Performance highlights {width="35%"}

### Row - Value boxes {height="30%"}

```{r}
#| label: DAX-valuebox-calcs
#| message: false
DAX_cum <- cum_perf |> filter(index == "DAX")
DAX_vol <- eu |> filter(index == "DAX") |> summarise(sd = sd(daily_return, na.rm = TRUE)) |> pull(sd)

DAX_best_month <- dax |>
group_by(month) |>
summarise(avg_daily_return = mean(daily_return, na.rm = TRUE), .groups = "drop") |>
arrange(desc(avg_daily_return)) |>
slice_head(n = 1)

DAX_worst_month <- dax |>
group_by(month) |>
summarise(avg_daily_return = mean(daily_return, na.rm = TRUE), .groups = "drop") |>
arrange(avg_daily_return) |>
slice_head(n = 1)

```

::: {.valuebox icon="percent" color="#31a354"}
DAX cumulative return:

`{r} percent(DAX_cum$cum_return, accuracy = 0.1)`
:::

::: {.valuebox icon="activity" color="#2b8cbe"}
DAX daily return volatility (sd):

`{r} percent(DAX_vol, accuracy = 0.01)`
:::

::: {.valuebox icon="star-fill" color="#d4af37"}
Best month (avg daily return):

`{r} format(DAX_best_month$month, "%Y-%m")`

`{r} percent(DAX_best_month$avg_daily_return, accuracy = 0.1)`
:::


### Row - Tabsets of tables {.tabset height="70%"}

::: {.card title="DAX — Top 30 daily returns"}
Best single-day returns.


```{r}
#| label: DAX-top-30-days
dax |>
filter(!is.na(daily_return)) |>
arrange(desc(daily_return)) |>
slice_head(n = 30) |>
transmute(
Date = date,
Price = round(price, 2),
`Daily Return` = percent(daily_return, accuracy = 0.01)
) |>
gt()

```

:::

::: {.card title="DAX — Bottom 30 daily returns"}
Worst single-day returns.

```{r}
#| label: DAX-bottom-30-days
dax |>
filter(!is.na(daily_return)) |>
arrange(daily_return) |>
slice_head(n = 30) |>
transmute(
Date = date,
Price = round(price, 2),
`Daily Return` = percent(daily_return, accuracy = 0.01)
) |>
gt()
```
:::


# 📈 FTSE Dashboard


## Column - Prices and returns over time {width="20%"}


```{r}
#| label: valuebox-generall-2
#| message: false
top_index <- cum_perf |> slice_head(n = 1)
bottom_index <- cum_perf |> slice_tail(n = 1)

dax_latest <- latest |> filter(index == "DAX")
ftse_latest <- latest |> filter(index == "FTSE")

```

::: {.valuebox icon="graph-up-arrow" color="#2b8cbe"}
Best cumulative performer:

`{r} as.character(top_index$index)`

`{r} percent(top_index$cum_return, accuracy = 0.1)`
:::


::: {.valuebox icon="graph-down-arrow" color="#d95f0e"}
Worst cumulative performer:

`{r} as.character(bottom_index$index)`

`{r} percent(bottom_index$cum_return, accuracy = 0.1)`
:::

::: {.valuebox icon="calendar-event" color="#756bb1"}
Latest observation date:

`{r} as.character(max(latest$latest_date))`
:::

## Column - Prices and returns over time {width="45%"}

### Row - Price level over time {height="60%"}

```{r}
#| label: ftse-price
#| title: FTSE — Price level
#| fig-width: 10
#| fig-asp: 0.618
ftse |>
ggplot(aes(x = date, y = price)) +
geom_line() +
scale_y_continuous(labels = label_number()) +
labs(x = NULL, y = NULL) +
theme_minimal()

```

### Row - Monthly average returns {height="40%"}

::: {.card title="FTSE — Monthly average returns"}
Average daily return within each month (approx).

```{r}
#| label: ftse-monthly-returns
#| fig-width: 10
#| fig-asp: 0.3
ftse |>
group_by(month) |>
summarise(avg_daily_return = mean(daily_return, na.rm = TRUE), .groups = "drop") |>
ggplot(aes(x = month, y = avg_daily_return)) +
geom_line() +
geom_point(size = 0.8) +
scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
scale_y_continuous(labels = label_percent(accuracy = 0.1)) +
labs(x = "Month", y = NULL) +
theme_minimal()

```

:::

## Column - Performance highlights {width="35%"}
### Row - Value boxes {height="30%"}

```{r}
#| label: ftse-valuebox-calcs
#| message: false
ftse_cum <- cum_perf |> filter(index == "FTSE")
ftse_vol <- eu |> filter(index == "FTSE") |> summarise(sd = sd(daily_return, na.rm = TRUE)) |> pull(sd)

ftse_best_month <- ftse |>
group_by(month) |>
summarise(avg_daily_return = mean(daily_return, na.rm = TRUE), .groups = "drop") |>
arrange(desc(avg_daily_return)) |>
slice_head(n = 1)

ftse_worst_month <- ftse |>
group_by(month) |>
summarise(avg_daily_return = mean(daily_return, na.rm = TRUE), .groups = "drop") |>
arrange(avg_daily_return) |>
slice_head(n = 1)

```

::: {.valuebox icon="percent" color="#31a354"}
FTSE cumulative return:

`{r} percent(ftse_cum$cum_return, accuracy = 0.1)`
:::

::: {.valuebox icon="activity" color="#2b8cbe"}
FTSE daily return volatility (sd):

`{r} percent(ftse_vol, accuracy = 0.01)`
:::

::: {.valuebox icon="star-fill" color="#d4af37"}
Best month (avg daily return):

`{r} format(ftse_best_month$month, "%Y-%m")`

`{r} percent(ftse_best_month$avg_daily_return, accuracy = 0.1)`
:::

### Row - Tabsets of tables {.tabset height="70%"}

::: {.card title="FTSE — Top 30 daily returns"}
Best single-day returns.


```{r}
#| label: ftse-top-30-days
ftse |>
filter(!is.na(daily_return)) |>
arrange(desc(daily_return)) |>
slice_head(n = 30) |>
transmute(
Date = date,
Price = round(price, 2),
`Daily Return` = percent(daily_return, accuracy = 0.01)
) |>
gt()

```

:::

::: {.card title="FTSE — Bottom 30 daily returns"}
Worst single-day returns.

```{r}
#| label: ftse-bottom-30-days
ftse |>
filter(!is.na(daily_return)) |>
arrange(daily_return) |>
slice_head(n = 30) |>
transmute(
Date = date,
Price = round(price, 2),
`Daily Return` = percent(daily_return, accuracy = 0.01)
) |>
gt()
```
:::

