---
title: DANL 310 Homework 1
subtitle: "`ggplot2` & `dplyr` Fundamentals  · Quarto Blogging"
author: Will Akin
date: last-modified
from: markdown+emoji
execute:
  eval: false
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 5
---

```{r}
#| include: false
library(tidyverse)
library(skimr)
library(dbplyr)
```


# 📌 Directions

- Submit **one Quarto document** (`.qmd`) to Brightspace:

  - `danl-310-hw1-LASTNAME-FIRSTNAME.qmd`  
    (e.g., `danl-310-hw1-choe-byeonghak.qmd`)

- **Due:** **February 18, 2026, 11:59 P.M. (ET)**

- For **visualization questions**, you must provide:
  1) the **`ggplot2` code**, and  
  2) a **written comment** (2–4 sentences) interpreting the corresponding figure.

- Unless a question says otherwise, use `dplyr` verbs (`filter()`, `select()`, `mutate()`, `group_by()`, `summarise()`, `arrange()`, `count()`, etc.) and `ggplot2`.

<br>

# ✅ Setup

```{r}
#| echo: true
library(tidyverse)
library(skimr)
```

---

<br><br>

# Part 1. Data Visualization & Summaries

## A. Orange Juice Promotions (`oj`)

Use the following dataset for Questions 1–6.

```{r}
#| echo: true
oj <- read_csv("https://bcdanl.github.io/data/dominick_oj_na.csv")
```

### Question 1. Quick inspection
1. Print the first 10 rows of `oj`.
2. Use `skimr::skim()` (or another method) to summarize the variables.

```{r}
head(oj, 10)
skimr::skim(oj)
```

<br>

### Question 2. Brand-level descriptive statistics (filter + group summary)
Filter to the three brands: **tropicana**, **minute.maid**, and **dominicks**. Then compute **mean** and **standard deviation** of:

- `sales`
- `price`

for each brand.

```{r}
oj |> filter(brand %in% c("tropicana", "minute.maid", "dominicks")) |>
  group_by(brand) |>
  summarise(    mean_sales = mean(sales, na.rm = TRUE),
    sd_sales   = sd(sales, na.rm = TRUE),
    mean_price = mean(price, na.rm = TRUE),
    sd_price   = sd(price, na.rm = TRUE)
  ) |>
  arrange(desc(mean_sales))
```

Write 2–3 sentences comparing the three brands based on your results.

# Dominicks has the greatest mean sales.
# Tropicana has the greatest mean price

<br>

### Question 3. Remove missing values
Create a new data frame `oj_no_na` that removes observations with **missing** values in **either** `price` or `sales`.

1. Show the number of rows in `oj` and `oj_no_na`.
2. Report how many rows were removed.

```{r}
oj_no_na <- oj |>
  filter(!is.na(price) & !is.na(sales))


nrow(oj)

nrow(oj_no_na)
```
# 6 rows were removed

<br>

### Question 4. Price distribution by brand (ggplot + comment)
Using `oj_no_na`, make a figure that compares the **distribution of `price` across `brand`**.

- Choose ONE main approach (e.g., boxplot, violin plot, density plot, histogram with facets).
- Include clear axis labels and a readable title.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(data = oj_no_na,
       mapping = 
         aes(x = brand,
             y = price)) + 
  labs(title = "Price Distribution by Brand") +
  geom_boxplot() 
```

**(b) Comment (2–3 sentences):** What differences do you see across brands (center, spread, skewness, outliers, etc.)?
# Dominicks has a few outliiers
# Dominicks has the lowest price median
<br>

### Question 5. Log–log price–sales relationship by brand (ggplot + comment)
Using `oj_no_na`, visualize how the relationship between:

- `log10(sales)` and `log10(price)`

varies by **brand**.

- Use a scatter plot with transparency (e.g., `alpha = 0.3`).
- Add a fitted line (e.g., `geom_smooth(method = "lm", se = FALSE)`), and
- Use faceting OR color to distinguish brands.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot( data = oj_no_na,
        mapping =
          aes(x = log10(price), 
              y = log10(sales),
              color = brand) ) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
    labs(
    title = "Log(Sales) vs Log(Price) by Brand")
```

**(b) Comment (2–3 sentences):** Do you see evidence that higher prices are associated with lower sales? Does the pattern differ by brand?
# Yes this graph conveys that higher prices are associated with lower sales, conveyed by Dominicks.  Tropicana conveys that they acheive majority of sales at a higher price.
<br>

### Question 6. Log–log relationship by brand and ad status (ggplot + comment)
Now extend Question 5 by incorporating `ad_status`. Visualize how the relationship between:

- `log10(sales)` and `log10(price)`

varies by **brand** *and* **`ad_status`**.

- Use faceting (`facet_grid()` or `facet_wrap()`), and/or
- Use color/shape for `ad_status`.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(oj_no_na,
       aes(x = log10(price),
           y = log10(sales),
           color = ad_status)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ brand) +
  labs(
    title = "Log(Sales) vs Log(Price) by Brand and Ad Status")
```

**(b) Comment (2–3 sentences):** How does advertising status appear to shift the relationship (level shift, slope change, or no clear difference)?

# With advertisment, brands achieve higher sales at lower prices.  The slope increased with the addition of advertisements.

---

## B. MLB Batting Trends (`mlb_bat`)

Use the following dataset for Questions 7–8.

```{r}
#| echo: true
mlb_bat <- read_csv("https://bcdanl.github.io/data/MLB_batting.csv")
```
```{r}
colnames(mlb_bat)

```

### Question 7. Create yearly hit percentages (data transformation)
Create a data frame `mlb_hit_pct` that contains **yearly hit percentages** for each `hit_type`
(**Single**, **Double**, **Triple**, **HomeRun**).

Your final table should have:

- `year`
- `hit_type`
- `hit_pct` (a percentage or proportion)


```{r}
mlb_hit_pct <- mlb_bat |>
  filter(hit_type %in% c("Single", "Double", "Triple", "HomeRun")) |>
  select(
    year,
    hit_type,
    hit_pct = percentage
  )

mlb_hit_pct
```

<br>

### Question 8. Visualize trends (ggplot + comment)
Make a figure that shows the yearly trends in hit percentages for each `hit_type`.

**(a) Provide your `ggplot2` code:**

```{r}
library(tidyverse)

ggplot(mlb_hit_pct, 
       aes(x = year, 
           y = hit_pct, 
           color = hit_type)) +
  geom_line(size = 1) +
  labs(
    title = "Yearly Trends in Hit Percentages by Hit Type")
```

**(b) Comment (2–3 sentences):** Which hit types are increasing/decreasing over time? Any notable breaks or eras?
# Double and HomeRun are increasing.
# Single and Triple are decreasing.
# No noticable breaks.
---

## C. Titanic Survival (`titanic`)

Use the following dataset for Questions 9–13.

```{r}
#| echo: true
titanic <- read_csv("https://bcdanl.github.io/data/titanic_cleaned.csv")
```

### Question 9. Two-way count table
Create `titanic_class_survival` that counts passengers by `class` and `survived`.

```{r}
titanic_class_survival <- titanic |>
  count(class, survived)

titanic_class_survival
```

<br>

### Question 10. Age distribution by class and gender (ggplot + comment)
Visualize how the **distribution of `age`** varies across `class` and `gender`.

**(a) Provide your `ggplot2` code:**


```{r}
ggplot( data = titanic,
        mapping = 
          aes(x = age) ) + 
  geom_histogram(bins = 20) +
  facet_grid(gender ~ class) +
  labs(
    title = "Age Distribution by Passenger Class and Gender")
```
<br>
**(b) Comment (2–3 sentences):** What differences do you see across class and gender?

# 3rd class majority is 30 year old aged men
# There is a small amount of 60 year olds in the 2nd and 3rd class.

### Question 11. Survival rate by class and gender (data + ggplot + comment)
1. Create a summary table with the **survival rate** (proportion survived) by `class` and `gender`.
2. Visualize the survival rates.

**(a) Provide your `dplyr` code for the summary table:**

```{r}
titanic_survival <- titanic |>
  group_by(class, gender) |>
  summarise(
    survival_rate = mean(survived, na.rm = TRUE),
    .groups = "drop"
  )

titanic_survival
```

**(b) Provide your `ggplot2` code:**

```{r}
ggplot(titanic_survival, 
       aes(x = class, 
           y = survival_rate, 
           color = gender)) +
geom_col(position = "dodge") +
  labs(
    title = "Titanic Survival Rate by Class and Gender")
```

**(c) Comment (2–3 sentences):** Which groups have the highest/lowest survival rates?
# Females overall have the greatest survival rate.
# Females in 1st class have the greatest survival rate.
<br>

### Question 12. Conditional distribution of survived (ggplot + comment)
Create a plot that shows the **distribution of `survived`** across `class` and `gender` using proportions (not raw counts).

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(titanic,
       aes(x = class,
           fill = survived)) +
  geom_bar(position = "fill") +
  facet_wrap(~ gender) +
  labs(
    title = "Conditional Distribution of Survival by Class and Gender")
```

**(b) Comment (2–3 sentences):** How does using proportions (instead of counts) change your interpretation?
# It makes it easier for the viewer to see the proportion of survivor/victims.
# The filled column additionally make it easier to decipher survival rates between gender.
<br>

### Question 13. Short interpretation
Write **3 bullet-point insights** supported by your tables/figures in Questions 9–12.
# Women have a higher chance of survival amongst gender.
# Women in the first and second class have the highest chance of survival.
# Men in the 2nd and 3rd class have a very similar (and lowest) chance of survival.
---

## D. NYC Dog Licenses (`nyc_dogs`)

Use the following dataset for Questions 14–16.

```{r}
#| echo: true
nyc_dogs <- read_csv("https://bcdanl.github.io/data/nyc_dogs_cleaned.csv")
```

### Question 14. Breed frequency table
Create `nyc_dogs_breeds` with:

- non-missing `breed`,
- `n >= 2000`, and
- sorted by `n` descending.

```{r}
nyc_dogs_breeds <- nyc_dogs |>
  filter(!is.na(breed)) |>        
  count(breed) |>                 
  filter(n >= 2000) |>            
  arrange(desc(n))                

nyc_dogs_breeds
```

<br>

### Question 15. Visualize the distribution of breeds (ggplot + comment)
Using `nyc_dogs_breeds`, make a figure that shows the distribution of breed counts.

**(a) Provide your `ggplot2` code:**

```{r}
ggplot(nyc_dogs_breeds, 
       aes(x = breed,
           y = n)) +
geom_col(position = "dodge") +
  coord_flip() + # beter vizualization
  labs(
    title = "Dog Breed Distribution")
```


**(b) Comment (2–3 sentences):** Is the distribution concentrated in a few breeds or spread out?
# The distribution is spread out with a few dominating breeds.
# The top breeds are yorkshier terrier and shih tzu.
<br>

### Question 16. “Top breeds” focus (ggplot + comment)
Create a plot of the **top 10 breeds** (by count) and comment on what you observe.

**(a) Provide your `dplyr` + `ggplot2` code:**

```{r}
top10_breeds <- nyc_dogs_breeds |>
  head(10)

ggplot(top10_breeds, 
       aes(x = breed,
           y = n)) +
geom_col(position = "dodge") +
  coord_flip() + # beter vizualization
  labs(
    title = "Dog Breed Distribution")
```

**(b) Comment (2–3 sentences):** Any surprises? What might explain the pattern?
# No surprises.
# The top breeds are still yorkshier terrier and shih tzu.
---

<br><br>

# Part 2. Data Transformation (`nyc_payroll_2025`)

For Questions 17–27, use `nyc_payroll_2025`.

For variable descriptions, see: Citywide Payroll Data (Fiscal Year) on NYC Open Data.
# Original loading data errors- used AI to load data
```{r}
library(readr)

# 1. Download the zip file
download.file(
  "https://bcdanl.github.io/data/nyc_payroll_2025.zip",
  destfile = "nyc_payroll_2025.zip",
  mode = "wb"
)

# 2. Unzip it
unzip("nyc_payroll_2025.zip")

# 3. Read the CSV file
nyc_payroll_2025 <- read_csv("nyc_payroll_2025.csv")

```

```{r}
# #| echo: true
# nyc_payroll_2025 <- read_csv("https://bcdanl.github.io/data/nyc_payroll_2025.zip")
```

### Question 17. Base salary by borough (filter + summarise)
Compute the **mean** and **standard deviation** of `Base_Salary` for workers whose `Work_Location_Borough` is:

- `"MANHATTAN"`
- `"QUEENS"`

Report the two means and two SDs in your write-up.

```{r}
nyc_payroll_2025 |> filter(Work_Location_Borough %in% c("MANHATTAN", "QUEENS")) |>
  group_by(Work_Location_Borough) |>
  summarise(    mean_sales = mean(Base_Salary, na.rm = TRUE),
    sd_Base_Salary   = sd(Base_Salary, na.rm = TRUE),
    mean_Base_Salary = mean(Base_Salary, na.rm = TRUE),
) |>
  arrange(desc(mean_Base_Salary))
```

<br>

### Question 18. High base salary filter
Filter the data to keep only records where `Base_Salary >= 100000`. Report how many rows remain.

```{r}
nyc_payroll_2025 |>
  filter(Base_Salary >= 100000) |>
nrow()
```

<br>

### Question 19. Distinct agency–title pairs
Select only **distinct** combinations of `Agency_Name` and `Title_Description`.

```{r}
nyc_payroll_2025 |> 
distinct(Agency_Name, Title_Description)
```

<br>

### Question 20. Top paid by regular gross pay
Arrange employees by `Regular_Gross_Paid` in **descending** order and show the **top 10** rows with:

- `First_Name`, `Last_Name`
- `Agency_Name`
- `Title_Description`
- `Regular_Gross_Paid`

```{r}
nyc_payroll_2025 |>
  arrange(desc(Regular_Gross_Paid)) |>
  select(
    First_Name,
    Last_Name,
    Agency_Name,
    Title_Description,
    Regular_Gross_Paid
  ) |>
  head(10)
```

<br>

### Question 21. Select + rename
Select `Title_Description` and rename it to `Title`. Also select `Agency_Name`, `First_Name`, `Last_Name`, and `Base_Salary`.

```{r}
nyc_payroll_2025 |> 
  rename(Title = Title_Description)
  
nyc_payroll_2025 |> 
  select(
    Agency_Name,
    First_Name,
    Last_Name,
    Base_Salary
  )
```

<br>


### Question 22. Create new pay variables (`mutate()`)
Use `mutate()` to create two new variables:

- `Total_Pay = Regular_Gross_Paid + Total_OT_Paid + Total_Other_Pay`
- `OT_Share = Total_OT_Paid / Total_Pay`

Then show the first 10 rows of:

- `First_Name`, `Last_Name`
- `Agency_Name`
- `Base_Salary`
- `Total_Pay`, `OT_Share`

```{r}
nyc_payroll_2025 |> 
  mutate(Total_Pay = Regular_Gross_Paid + Total_OT_Paid + Total_Other_Pay) |> 
  mutate(OT_Share = Total_OT_Paid / Total_Pay) |>
  select(
    First_Name,
    Last_Name,
    Agency_Name,
    Base_Salary,
    Total_Pay,
    OT_Share
  ) |>
head(10)
```

<br>

### Question 23. Borough pay summary (`group_by()` + `summarise()`)
Using the variables you created in **Question 22**, group the data by `Work_Location_Borough` and compute:

- number of employees `n`
- mean `Base_Salary`
- mean `Total_Pay`
- mean `OT_Share` (ignore missing values)

Arrange the results by mean `Total_Pay` in **descending** order and show the summary table.

```{r}
nyc_payroll_2025 |>
  mutate(
    Total_Pay = Regular_Gross_Paid + Total_OT_Paid + Total_Other_Pay,
    OT_Share  = Total_OT_Paid / Total_Pay
  ) |>
  group_by(Work_Location_Borough) |>
  summarise(
    n = n(),
    mean_Base_Salary = mean(Base_Salary, na.rm = TRUE),
    mean_Total_Pay   = mean(Total_Pay, na.rm = TRUE),
    mean_OT_Share    = mean(OT_Share, na.rm = TRUE)
  ) |>
  arrange(desc(mean_Total_Pay))
```

<br>

### Question 24. Police Department overtime
Filter to `Agency_Name == "POLICE DEPARTMENT"` and arrange by `Total_OT_Paid`.

1. Show the **10 smallest** overtime values.
2. Show the **10 largest** overtime values.

```{r}
nyc_payroll_2025 |> 
  filter(Agency_Name == "POLICE DEPARTMENT") |> 
  arrange(Total_OT_Paid) |> 
head(10)

nyc_payroll_2025 |> 
  filter(Agency_Name == "POLICE DEPARTMENT") |> 
  arrange(desc(Total_OT_Paid)) |> 
head(10)
```

<br>

### Question 25. Per annum employees
Filter to `Pay_Basis == "per Annum"` and select only:

- `First_Name`, `Last_Name`, `Base_Salary`

```{r}
nyc_payroll_2025 |> 
  filter(Pay_Basis == "per Annum") |> 
  select(
    First_Name,
    Last_Name,
    Base_Salary
  )
```

<br>

### Question 26. Borough then salary
Arrange by `Work_Location_Borough` (ascending) and then `Base_Salary` (descending). Then show the first 15 rows.

```{r}
nyc_payroll_2025 |> 
  arrange(Work_Location_Borough) |> 
  arrange(desc(Base_Salary)) |> 
  head(15)
```

<br>

### Question 27. Remove missing last names + count
Remove observations where `Last_Name` is missing (`NA`). Then report the remaining number of rows.

```{r}
nyc_payroll_2025 |> 
  filter(!is.na(Last_Name)) |> 
  nrow()
```

---

<br><br>

# Part 3. Quarto Blogging (`ice_cream`)

Include the URL to your published blog post.

**Answer:**

file:///Users/willakin/Documents/willakin16.github.io/posts/DANL310%20ICECREAM/danl310icecream_df.html
