---
title: "DANL 310 Mental Health Analysis Project"
subtitle: "DANL 310: Data Visualization and Presentation"
author:
  - name: Will Akin
    affiliation: SUNY Geneseo
date: last-modified

execute:
  echo: true
  eval: true
  warning: false
  message: false
  fig-width: 9
  fig-height: 6

format:
  html:
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-expand: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: atom-one
---

```{r setup}
#| include: false
library(tidyverse)
library(skimr)
library(hrbrthemes)
library(rmarkdown)

# Add other packages as needed.
# Examples:
# library(lubridate)
# library(janitor)
# library(scales)
# library(viridis)
# library(plotly)
# library(DT)
# library(sf)

# Feel free to revise this theme.
theme_set(
  theme_ipsum() +
    theme(
      strip.background = element_rect(fill = "lightgray"),
      axis.title.x = element_text(size = rel(1.33)),
      axis.title.y = element_text(size = rel(1.33))
    )
)

# Optional color-blind-friendly scales.
# scale_colour_discrete <- function(...) scale_colour_viridis_d(...)
# scale_fill_discrete <- function(...) scale_fill_viridis_d(...)
```

# Big Idea

As a data analytics and sociomedical science major, I am interested in studying public health datasets to better understand mental health trends and the factors associated with disorders such as anxiety and depression. This topic is especially important because mental health disorders affect millions of people and can significantly impact overall well-being and quality of life. By analyzing these patterns and trends, public health professionals, healthcare providers, policymakers, and researchers can develop more effective prevention strategies, interventions, and support systems for affected individuals and communities.

# Guiding Question or Story Theme

How can public health data analysis help identify the social, behavioral, and demographic factors that contribute to mental health disorders such as anxiety and depression?

# Background and Context

Mental health data is important because it allows researchers to identify social, behavioral, and demographic variables that may increase a person’s risk for mental health conditions.

# Data

```{r data-import}
library(tidyverse)
df <- read_csv("/Users/willakin/Documents/DANL 310/DANL 310/DANL 310/anxiety_depression_data.csv")
```

## Key Variables

```{r}
colnames(df)
```
```{r}
library(skimr)
skim(df)
```
```{r}
summary(df)
```


## Unit of Observation

Each row in the dataset represents one individual participant and their demographic, lifestyle, social, and mental health information. The variables include characteristics such as age, gender, education level, employment status, sleep hours, physical activity, stress levels, social support, and mental health measures including anxiety and depression scores. Together, each row provides a profile of a person that can be analyzed to identify patterns and factors associated with mental health outcomes.

## Data Source

The data set I used is called Anxiety and Depression Mental Health Factors. I retrieved this data from Kaggle. I accessed this data on April 20th. This dataset contains related to anxiety, depression, and mental health influences. It includes demographic details, lifestyle habits, mental health indicators, medical history, coping mechanisms, and stress factors. The dataset is designed for mental health analysis, predictive modeling, and research on the impact of various factors on mental well-being. The link is https://www.kaggle.com/datasets/ak0212/anxiety-and-depression-mental-health-factors

# Data Preparation and Transformation

First I refactored the data variables. Then I created Age group variables that divided the age variables into different age ranges for better analysis. I also reshaped the data frame to Outcome and Score for a better analysis on how the variables of Score impact the variables of Outcome.


```{r data-preparation}
df <- df %>%
  mutate(
    Gender = as.factor(Gender),
    Education_Level = as.factor(Education_Level),
    Employment_Status = as.factor(Employment_Status),
    Family_History_Mental_Illness = as.factor(Family_History_Mental_Illness),
    Chronic_Illnesses = as.factor(Chronic_Illnesses),
    Medication_Use = as.factor(Medication_Use),
    Therapy = as.factor(Therapy),
    Meditation = as.factor(Meditation),
    Substance_Use = as.factor(Substance_Use)
  )
```

```{r}
df <- df %>%
  mutate(
    Age_Group = case_when(
      Age < 18 ~ "Under 18",
      Age >= 18 & Age <= 24 ~ "18–24",
      Age >= 25 & Age <= 34 ~ "25–34",
      Age >= 35 & Age <= 44 ~ "35–44",
      Age >= 45 & Age <= 54 ~ "45–54",
      Age >= 55 ~ "55+"
    )
  )

```


```{r}

df_long <- df %>%
  pivot_longer(
    cols = c(Anxiety_Score, Depression_Score, Stress_Level),
    names_to = "Outcome",
    values_to = "Score"
  )

```


# Descriptive Statistics

```{r}
library(tidyverse)

summary_table <- df %>%
  summarise(
    Anxiety_Mean = mean(Anxiety_Score, na.rm = TRUE),
    Anxiety_SD = sd(Anxiety_Score, na.rm = TRUE),

    Depression_Mean = mean(Depression_Score, na.rm = TRUE),
    Depression_SD = sd(Depression_Score, na.rm = TRUE),

    Stress_Mean = mean(Stress_Level, na.rm = TRUE),
    Stress_SD = sd(Stress_Level, na.rm = TRUE)
  )
summary_table
```
The mean values of Anxiety, Depression, and Stress Level can be used as benchmarks when comparing the levels across different variables.

## Basic Summary

```{r summary-statistics}
skim(df)
```

## Initial Observations

Briefly explain what the descriptive statistics show.

Focus on patterns such as:

- largest or smallest groups,
- typical values,
- variation across groups,
- outliers or unusual cases,
- missing data issues, and
- patterns that set up your main story.

# Data Visualization and Storytelling

## Visualization 1: Overall Distribution

```{r visualization-1}
library(tidyverse)
library(viridis)

ggplot(df_long, aes(x = Score, fill = after_stat(count))) +
  geom_histogram(bins = 15, color = "white") +
  facet_wrap(~ Outcome, scales = "free_x") +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "Distribution of Mental Health Measures",
    x = "Score",
    y = "Count",
    fill = "Frequency"
  )
```

**Interpretation:**  
Interpretation: This graph shows the overall distribution of anxiety, depression, and stress level scores across the data set. This allows the audience to see how often each score for each of the tested variables are observed.

## Visualization 2: Mental Health Scores by Education Level

```{r visualization-2}
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)

ggplot(df_long, aes(x = Education_Level, y = Score, fill = Education_Level)) +
  geom_boxplot() +
  facet_wrap(~ Outcome) +
  scale_fill_viridis_d()+
  theme_minimal() +
  labs(
    title = "Mental Health Scores by Education Level",
    x = "Education Level",
    y = "Score"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**Interpretation:**  
Interpretation: Mean Stress Level has little differentiation per education level.
The Master’s education Level had a  low mean anxiety score, but a high depression score. 


## Visualization 3: Mental Health Outcomes by Age Group

```{r visualization-3}
library(tidyverse)
library(viridis)

df_long_summary <- df_long %>%
  group_by(Age_Group, Outcome) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop")

ggplot(df_long_summary, aes(x = Age_Group, y = mean_score, fill = Age_Group)) +
  geom_col() +
  scale_fill_viridis_d() +
  facet_wrap(~Outcome, scales = "free_y") +
  labs(
    title = "Average Mental Health Outcomes by Age Group",
    x = "Age Group",
    y = "Mean Score"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

```

**Interpretation:**  
Interpretation: 18-24 year olds had the greatest anxiety and depression score, possibly due to school. 35-44 year olds had the greatest stress levels.

## Visualization 4: Mental Health Scores by Gender

```{r visualization-4}
ggplot(df_long, aes(x = Gender, y = Score, fill = Gender)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ Outcome) +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(
    title = "Mental Health Scores by Gender",
    x = "Gender",
    y = "Score"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

**Interpretation:**  
Other Gender had greatest Depression and Stress Level Scores, possibly due to social stigma.


## Shiny App

```{=html}
<div style="display: flex; justify-content: center; margin: 1em 0;">
  <iframe
    src="YOUR_SHINY_APP_URL"
    width="800"
    height="800"
    frameborder="0"
    style="border: none;"
  ></iframe>
</div>
```

The Shiny app conveys the distribution of variables by Age Group.

## Dashboard

- **Dashboard:** [Insert link here](https://willakin16.github.io/DANL310_Dashboard.html)

This dashboard clearly conveys the visualizations in a cleaner matter, where all graphs can be crossed-analyzed.


# Connecting the Dots

The descriptive statistics, data transformations, and visualizations all work together to answer the guiding question by showing how demographic, behavioral, and social factors are connected to mental health outcomes such as anxiety, depression, and stress. The transformations made the data easier to analyze by grouping ages into meaningful categories and reshaping the dataset into a format that allowed direct comparison between different mental health measures. This helped reveal patterns across groups and improved the clarity of the visualizations.

One of the strongest patterns in the data was the relationship between age and mental health outcomes. The visualizations showed that individuals ages 18–24 had the highest average anxiety and depression scores, while individuals ages 35–44 reported the highest stress levels. These findings suggest that different stages of life may expose people to different stressors, such as academic pressure for younger adults and career or family responsibilities for middle-aged adults. This demonstrates how demographic characteristics can influence mental health in different ways.

Another important finding was the variation in mental health outcomes across gender categories. Participants identifying as “Other” showed higher average depression and stress scores than other groups. This result highlights the possible impact of social stigma, discrimination, and unequal access to mental health support on marginalized populations. The education-level analysis also revealed differences in mental health patterns, suggesting that educational experiences and expectations may affect anxiety, stress, and depression differently.

The descriptive statistics supported these visual findings by summarizing the overall levels and variability of anxiety, depression, and stress within the dataset. Together, the statistics and visualizations provide evidence that mental health is influenced by multiple interconnected social and behavioral factors rather than a single cause.

This analysis is socially and publicly relevant because mental health disorders affect millions of people and are closely tied to quality of life, productivity, relationships, and physical health. By identifying which groups may experience higher levels of anxiety, depression, or stress, public health professionals and policymakers can design more targeted prevention programs, mental health resources, and intervention strategies. Understanding these patterns can help improve mental health awareness, reduce stigma, and support communities that may be at greater risk.

# Discussion

The data suggest that mental health outcomes vary across demographic and social groups. Younger adults between the ages of 18–24 showed higher average anxiety and depression scores compared to other age groups. One possible explanation is that people in this age range may experience stress related to school, career uncertainty, finances, or major life transitions. The pattern is consistent with concerns often discussed in public health research about mental health challenges among young adults.

The comparison across age groups also showed that adults ages 35–44 had the highest stress levels. This trend raises questions about how work responsibilities, financial pressures, and family obligations may influence stress during middle adulthood. While the data cannot prove these factors directly cause stress, the pattern suggests that life stage differences may be associated with different mental health experiences.

The gender analysis showed that participants categorized as “Other” had higher average depression and stress scores than other gender groups. This comparison suggests that gender minority groups may experience additional social or emotional challenges. One possible explanation is that experiences such as discrimination, stigma, or reduced social support could be connected to these patterns. However, the analysis only identifies associations within the dataset and does not establish causation.

The education-level comparisons showed relatively small differences in stress levels across groups, although some variation appeared in anxiety and depression scores. For example, participants with Master’s degrees showed lower average anxiety scores but somewhat higher depression scores. This pattern suggests that education may relate differently to different aspects of mental health, and it raises questions about how academic expectations, career pressure, or work-life balance may affect emotional well-being.

Overall, the findings suggest that mental health outcomes are associated with a combination of demographic, social, and behavioral factors rather than a single influence. The analysis highlights patterns that may help public health professionals identify groups that could benefit from additional mental health support, outreach, or prevention programs.

## Limitations

Although this project provides useful insight into mental health patterns, there are several limitations that should be considered when interpreting the findings.

First, the dataset may not fully represent all populations. Because the data came from a Kaggle dataset rather than a nationally representative survey, the participants may not reflect the experiences of all demographic groups, geographic regions, or socioeconomic backgrounds. This means the findings may not generalize to the broader population.

Another limitation is that the analysis is descriptive rather than causal. The visualizations and summary statistics can identify patterns and associations between variables, but they cannot prove that one factor directly causes anxiety, depression, or stress.

The dataset may also contain measurement limitations. Mental health variables such as anxiety, depression, and stress are often based on self-reported responses, which can be influenced by personal interpretation, honesty, or differences in how participants understand survey questions. This means the scores may not perfectly capture an individual’s true mental health condition.

In addition, some potentially important variables that also influence anxiety depression, and stress level may be missing from the dataset.

# Conclusion

This project explored how demographic, social, and behavioral factors are associated with mental health outcomes such as anxiety, depression, and stress. The analysis revealed several important patterns in the data, including higher anxiety and depression scores among younger adults ages 18–24, higher stress levels among adults ages 35–44, and elevated depression and stress scores among participants identifying as “Other” in gender categories. The visualizations and descriptive statistics suggest that mental health experiences differ across groups and may be connected to factors such as life stage, social pressures, education, and support systems.

These findings matter because mental health disorders affect millions of people and have major impacts on overall well-being, relationships, education, and workplace performance. While the analysis does not prove causation, it highlights patterns that can help public health professionals, researchers, and policymakers better understand which groups may be more vulnerable to mental health challenges. The main takeaway from this project is that mental health is shaped by a combination of interconnected factors, and data analysis can play an important role in identifying trends, raising awareness, and supporting more targeted mental health interventions and resources.

# Future Directions

If this project were expanded further, I would explore additional variables that may influence mental health outcomes, such as income level, access to healthcare, relationship status, or geographic location. Including more detailed social and economic variables could provide a broader understanding of the factors associated with anxiety, depression, and stress.

I would also be interested in analyzing data across a longer time period to examine how mental health trends change over time or in response to major events such as economic shifts or public health crises. In future work, predictive modeling techniques could also be explored to identify which combinations of factors are most strongly associated with mental health outcomes while still recognizing that correlation does not necessarily imply causation.

# Project Write Up

# References

Kaggle 4/20 - https://www.kaggle.com
